\documentclass[11pt,letterpaper]{article}

\usepackage[margin=1.9cm]{geometry}
\usepackage{tgtermes}
\usepackage{fancyhdr}

\pagestyle{fancy}
\fancyhf{}
\rhead{Errington, Jacob Thomas}
\lhead{LEADERSHIP \& INVOLVEMENT}
\cfoot{\thepage}

\begin{document}

\begin{itemize}
  \item
    In the 2015-2016 academic year, I was a part of the Hackboard, a selective
    organization of involved students who organize the annual instalment of
    McHacks, McGill's official hackathon.

  \item
    Furthermore, my duties on the hackboard included organizing and hosting
    community events throughout the year called hacknights. I frequently
    presented my personal projects at hacknights, and sometimes gave more
    involved talks, e.g. about version control with Git.

  \item
    I presented a paper by Simon Gravel titled \textit{The existence and
    abundance of ghost ancestors in biparental populations} to the First Choice
    Science students at Dawson College as an invited speaker.

  \item
    In fall 2016 for a graduate-level course on program analyses and
    transformations (COMP 621), I presented the paper \textit{Implementing lazy
    functional languages on stock hardware: the Spineless Tagless G-machine} by
    Simon Peyton Jones. I also presented my term project, an embedded
    domain-specific language for expressing analyses and transformations of an
    imperative language.

  \item
    I gave an introductory talk on dependently-typed programming in Idris at
    the 2017 Seminars in Undergraduate Mathematics in Montreal (SUMM).

  \item
    At the end of my internship at 1010data in New York City, I presented my
    work in a 20-minute talk to the CEO, the CTO, the chief scientist, and
    numerous other colleagues.

  \item
    I am currently one of two codirectors of technical initiatives at
    Building21, a new space at McGill for students, faculty, and staff where
    science and art come together. My duties include the maintenance of the
    computer infrastructure of the space and the planning of technical
    workshops.

  \item
    I am currently a teaching assistant for an introductory course at McGill
    (COMP 302) on functional programming. My duties include holding weekly
    office hours to assist students as well as grading assignments.
\end{itemize}

\end{document}
